A Pandemic-Scale Ancestral Recombination Graph for SARS-CoV-2

### 2025-11-04 latex/main.tex
Various minor updates

\documentclass[12pt,letterpaper]{article}
\usepackage[a4paper, total={7in, 10in}]{geometry}
\renewcommand{\familydefault}{\sfdefault}
\usepackage{graphicx}
\usepackage{helvet}
\usepackage{authblk}
\usepackage{hyperref}
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{orcidlink}
\usepackage[super,comma,sort&compress]
   {natbib}\bibliographystyle{numbered}
\usepackage[right]{lineno} \linenumbers

@@

% NOTE: Trying to clarify that this of the pandemic period and
% we're not trying to cover *all* history.
The ARG is a comprehensive and deeply validated record of
the pandemic phase of SARS-CoV-2 evolutionary history,
accurately reflecting a range of results from the literature.
First, we compared the sc2ts ARG
with the state-of-the-art UShER tree inferred from the same dataset
by pruning both to their intersection of
2,475,418 samples and 27,507 sites (STAR Methods).
% JK: this feels a bit wishy washy, but there really is no simple answer to
% this and I don't want to claim we're more parsimonious.
We found that they were very similar in terms of parsimony
(Document~\ref{sec:supplement_sc2ts_usher_parsimony})
and imputation of missing and ambiguous bases
(Document~\ref{sec:supplement_sc2ts_usher_imputation}).
We also compared the mutational spectra of major VOCs in the ARG
against the UShER tree and a previous study~\cite{Bloom2023} (STAR Methods),
finding close agreement (Figure~\ref{fig:mutational_spectra}).
We validated that the broad phylogenetic structure of
the ARG and UShER tree closely match by generating tanglegrams
for a set of representative samples from 814 well-sampled lineages
(STAR Methods; Figures~\ref{fig:tanglegram_base},
\ref{fig:tanglegram_delta},
\ref{fig:tanglegram_ba2},
\ref{fig:tanglegram_ba5}).
We also evaluated how well the ARG reflects the structure of
the Pango nomenclature system
by computing Pango assignments for all 2.75 million nodes in the ARG
using Pangolin~\cite{OToole2021assignment} (STAR Methods).
All but 21,680 of the exported sample alignments (0.87\%) agreed
with the source consensus sequences in terms of Pango lineage assignment
(comparable to the 0.27\% disagreement between pangolin-data
versions 1.21 and 1.29 in the source Viridian metadata).
A majority of these lineages (1473 of 2058)
have monophyletic origins in the ARG,
increasing to 1779 (86\%) when we allow
for multiple sibling origination nodes
(Document~\ref{sec:supplement_pango_lineages}).
The ARG is time-resolved, with internal node dates estimated
by a non-parametric method~\cite{Pope2025} (STAR Methods).
We show that the dates agree reasonably well with
NextStrain estimates (Figure~\ref{fig:pango_vs_nextstrain_node_times})
and overall accuracy is comparable to 
Chronumental~\cite{sanderson2021chronumental} applied to 
an UShER tree (Document~\ref{sec:evaluating_node_dates}).
Finally, we verified
that the origins of the
Alpha (Figure~\ref{fig:subgraph_alpha}, Document~\ref{sec:dd_alpha}),
Delta (Figure~\ref{fig:subgraph_delta}, Document~\ref{sec:dd_delta})
and
Omicron (Figure~\ref{fig:subgraph_omicron}, Document~\ref{sec:dd_omicron})
VOCs reflect expectations from the literature~\cite{RambautAlpha2020,ECDCreport,
Cherian2021,Farinholt2021,Stern2021,Dor2025,RambautOmicron2021}.

@@

We validated the recombination events in the sc2ts ARG
by running 3SEQ~\cite{Lam2018improved}
on the putative recombinant sequence trios (STAR Methods),
and 484 of the 855 (57\%) were classified as recombinant.
Validation status correlates strongly with our QC filter,
with 95\% of the QC-passing recombinants
and 29\% of the QC-failing recombinants
being classified as recombinant.
In addition to 3SEQ, we ran the CovRecomb~\cite{Li2024CovRecomb} 
and rebar recombination detection methods
on the putative recombinant sequences,
summarized in Table~\ref{tab:recombinant_validation}.
CovRecomb has a substantially lower validation rate
at 25\% among the QC-passing recombinants.
This is expected, as CovRecomb is specifically intended
to detect interlineage recombinants
and has previously been shown to be
less sensitive than other methods~\cite{Li2024CovRecomb}.
Rebar validation is intermediate, at 57\% among QC-passing recombinants.
(See also Figures~\ref{fig:recombinant_evidence_cov_recomb},
and \ref{fig:recombinant_evidence_rebar}.)
% We don't talk about RIPPLES results yet as we haven't discussed it, a
% and that would be confusing.

@@

Most of the deletions in these tables
occur in coding regions
(except the 1-base deletions at 28,271) and
do not cause frame shifts, involving 3, 6, or 9 sites
(except the 4-base deletion and the two 5-base deletions,
which are likely artefacts; see below).
All the deletions in these tables
(besides the artefactual cases)
have been linked to the emergence of major VOCs
(Alpha \cite{RambautAlpha2020},
Delta \cite{Stern2021},
Omicron B.1.1.529 \cite{RambautOmicron2021},
BA.1 \cite{RambautOmicron2021},
BA.2 \cite{RambautOmicron2021}, and
BA.4 \cite{Tegally2022})
and/or reported to be prevalent
in previous studies of deletions \cite{Li2023indels}.
All these major deletion events
(again, besides the artefactual cases)
are associated with the origins of major VOCs in the ARG
(see the Alpha subgraph in Figure~\ref{fig:subgraph_alpha};
the Delta subgraph in Figure~\ref{fig:subgraph_delta}; and
the Omicron subgraph in Figure~\ref{fig:subgraph_omicron}),
except the 1-base deletion at 28,271 on node 1436808
and the 9-base deletion at 686 on the BA.4-labeled node 2698748
(Table~\ref{tab:major_dels}).
Most of these deletions have been identified
as characteristic mutations of the major VOCs
in the studies or analysis cited in Table~\ref{tab:major_dels}.
Notably, eight of the major deletion events occur
in the N-terminal domain of the S1 unit of Spike
(a known hotspot of recurrent deletions \cite{McCarthy2021}
that plays a substantial role in antigenicity \cite{Harvey2021}),
which are all highly recurrent in the ARG
(Table~\ref{tab:rec_dels}).
These include known recurrent deletions:
the 6-base deletion at 21,765 and
the 3-base deletion at 21,991,
which are associated with
increased efficiency in cell entry \cite{Meng2021} and
antibody escape \cite{McCarthy2021}, respectively.
In addition, we found recurrent deletions in ORF8 and NSP3,
which have been suggested to be indel hotspots
[FIXME broken reference]
\cite{Akaishi2022}.
Together, these findings demonstrate that
the ARG captures genuine deletion events in SARS-CoV-2 evolution.
A comprehensive evolutionary analysis of the deletion events in the ARG
is beyond the scope of the current study;
however, we believe that the ARG, as a resource,
will enable and facilitate future studies
to better understand the role of deletions in SARS-CoV-2 evolution
(alongside recombination and single nucleotide mutations),
which to-date has been understudied in great part due to
a lack of suitable data models, methods, and tools.

@@

The B.1.1.7 (Alpha) variant emerged in the UK as the first major VOC.
Its origin is well reconstructed in the ARG,
with Figure \ref{fig:subgraph_alpha}
showing it to be a saltational lineage
characterized by a long branch preceding its origin.
Most of the characteristic Alpha mutations \cite{RambautAlpha2020}
(20 out of 23) in the ARG
occur immediately above the Alpha root.
However, a characteristic three-base correlated mutation
is instead associated with the parent node (labeled B.1.1) of the Alpha root,
suggesting that this mutation was acquired prior to the emergence of Alpha.
This parent node has an additional 5 children
(leading to 12 samples),
of which we display one (ERR4413600)
with no associated mutations,
meaning that it has an identical sequence
to the inferred ancestor of Alpha.
Below the root of Alpha, a single sample (ERR5178290)
appears without the characteristic mutation C3267T
(or any additional mutations).
However, this is only a single sample and
given that it also appears two months after the first Alpha samples
we treat its placement with caution.
Note that the list of mutations in \cite{RambautAlpha2020}
includes a T26801C in the M gene
which should no longer be considered characteristic of Alpha,
as the Wuhan-Hu-1 reference at 26,801 has since been updated to C.
In the ARG, a one-base non-coding deletion (position 28,271)
is also associated with the Alpha root;
although this is not in the original list of
characteristic Alpha mutations,
it has been suggested elsewhere as an Alpha-defining mutation 
(\cite{Li2023indels}).

@@

Delta belongs to the ``B.1.617'' clade,
which also includes Kappa (B.1.617.1) and B.1.617.3 \cite{ECDCreport}.
Previous phylogenetic analyses suggest that
Delta and B.1.617.3 are more closely related to each other
than either is to Kappa (\cite{Cherian2021,Farinholt2021}).
In the ARG, however, Delta and Kappa appear as sibling lineages
(see the subgraph in Figure~\ref{fig:subgraph_delta}).
This reflects the closest approximation to the relationships
suggested in the previous studies that could be inferred,
given that no B.1.617.3 samples passed our QC filters and
no seed B.1.617.3 samples were specified for ARG inference.

@@

\subsection{Additional analyses of the Pango X lineages}
\label{sec:supplement_indepth_pango_x_lineages}

\newcommand{\scghdiscussion}[1]{%
See \href{https://github.com/jeromekelleher/sc2ts-paper/discussions/#1}%
{github:sc2ts-paper/discussions/#1} for further details.}
\newcommand{\scghdiscussiontwo}[2]{%
See \href{https://github.com/jeromekelleher/sc2ts-paper/discussions/#1}%
{github:sc2ts-paper/discussions/#1} for further details on #2.}
\subsubsection{XA}
% mneumonic: deep dive XA
\label{sec:dd_xa}
XA traces to a well supported recombination node, with no reversions.
Note that the ``causal'' sample ERR5308556
(which triggers the initial recombination event)
is identical to the recombination node, and
therefore lacks C8090T,
which is one of the XA consensus mutations
(identified as those shared by over 90\% of XA samples).
As mentioned elsewhere,
the sc2ts detection results in terms of
the Pango lineages of the parents and the breakpoint intervals
agree with the results reported by Jackson et al. (\ref{tab:jackson}).
Two of the Jackson et al. XA samples
are present in the ARG (ERR5308556 and ERR5414941);
these samples are specifically marked in Figure~\ref{fig:subgraphs},
which is a summary of this subgraph.

@@

The displayed subgraph shows a small selection of the 192 XB samples.
XB was originally defined as a recombinant between
the closely related lineages B.1.634 (left parent) and B.1.631 (right parent)
from North and Central America \cite{Gutierrez2022}.
However, these parent lineages were not added to the ARG due to a high HMM cost.
Together with other lineages such as B.1.627
(sister to B.1.631 \cite{Gutierrez2022}),
they are likely to attach somewhere along the long branch
in the subgraph visualization (with $> 30$ mutations),
which also contains the key ``R2'' deletion
associated with XB / B.1.634 \cite{Gutierrez2022}.

We hypothesize that there are a set of samples
from a number of Pango lineages (including B.1.627, B.1.631, B.1.634)
which, should they be added to the ARG, would break up the long branch,
with XB being a recombination between two of them.
In the absence of these, sc2ts cannot detect XB as a recombinant.
Also, note that the XB consensus mutations all occur on or above the long branch.
We suspect that the child node of the long branch
represents the (potentially recombinant) origin of XB, and
the B.1 samples at the bottom left of the subgraph
may be misclassified by pangolin.

Further work is required to determine
why these additional samples were not gradually added
to the ARG by the daily sample matching process or
retrospective sample grouping.

>>>>

\documentclass[12pt,letterpaper]{article}
\usepackage[a4paper, total={7in, 10in}]{geometry}
\renewcommand{\familydefault}{\sfdefault}
\usepackage{graphicx}
\usepackage{helvet}
\usepackage{authblk}
\usepackage{hyperref}
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{orcidlink}
\usepackage[super,comma,sort&compress]
   {natbib}\bibliographystyle{numbered}
\usepackage[right]{lineno} \linenumbers
\usepackage[normalem]{ulem}

@@

% NOTE: Trying to clarify that this of the pandemic period and
% we're not trying to cover *all* history.
The ARG is a comprehensive and deeply validated record of
the pandemic phase of SARS-CoV-2 evolutionary history,
accurately reflecting a range of results from the literature.
First, we compared the sc2ts ARG
with the state-of-the-art UShER tree inferred from the same dataset
by pruning both to their intersection of
2,475,418 samples and 27,507 sites (STAR Methods).
% JK: this feels a bit wishy washy, but there really is no simple answer to
% this and I don't want to claim we're more parsimonious.
Eliminating recombination nodes and their descendants (which we expected to differ), we found very similar relationships among  lineages in terms of parsimony
(Document~\ref{sec:supplement_sc2ts_usher_parsimony})
and imputation of missing and ambiguous bases
(Document~\ref{sec:supplement_sc2ts_usher_imputation}).
We also compared the mutational spectra of major VOCs in the ARG
against the UShER tree and a previous study~\cite{Bloom2023} (STAR Methods),
finding close agreement (Figure~\ref{fig:mutational_spectra}).
We validated that the broad phylogenetic structure of
the ARG and UShER tree closely match by generating tanglegrams
for a set of representative samples from 814 well-sampled lineages
(STAR Methods; Figures~\ref{fig:tanglegram_base},
\ref{fig:tanglegram_delta},
\ref{fig:tanglegram_ba2},
\ref{fig:tanglegram_ba5}).
We also evaluated how well the ARG reflects the structure of
the Pango nomenclature system
by computing Pango assignments for all 2.75 million nodes in the ARG
using Pangolin~\cite{OToole2021assignment} (STAR Methods).
All but 21,680 of the exported sample alignments (0.87\%) agreed
with the source consensus sequences in terms of Pango lineage assignment
(comparable to the 0.27\% disagreement between pangolin-data
versions 1.21 and 1.29 in the source Viridian metadata).
A majority of these lineages (1473 of 2058)
have monophyletic origins in the ARG,
increasing to 1779 (86\%) when we allow
for multiple sibling origination nodes
(Document~\ref{sec:supplement_pango_lineages}).
The ARG is time-resolved, with internal node dates estimated
by a non-parametric method~\cite{Pope2025} (STAR Methods).
We show that the dates agree reasonably well with
NextStrain estimates (Figure~\ref{fig:pango_vs_nextstrain_node_times})
and overall accuracy is comparable to 
Chronumental~\cite{sanderson2021chronumental} applied to 
an UShER tree (Document~\ref{sec:evaluating_node_dates}).
Finally, we verified
that the origins of the
Alpha (Figure~\ref{fig:subgraph_alpha}, Document~\ref{sec:dd_alpha}),
Delta (Figure~\ref{fig:subgraph_delta}, Document~\ref{sec:dd_delta})
and
Omicron (Figure~\ref{fig:subgraph_omicron}, Document~\ref{sec:dd_omicron})
VOCs reflect expectations from the literature~\cite{RambautAlpha2020,ECDCreport,
Cherian2021,Farinholt2021,Stern2021,Dor2025,RambautOmicron2021}.

@@

We validated the recombination events in the sc2ts ARG
by running 3SEQ~\cite{Lam2018improved}
on the putative recombinant sequence trios (STAR Methods)\sout{,
and 484 of the 855 (57\%) were classified as recombinant.
Validation status correlates strongly with our QC filter,
with 95\% of the QC-passing recombinants
and 29\% of the QC-failing recombinants
being classified as recombinant.}
.
Of the 855 recombination events identified by sc2ts, 3SEQ verified 57\% of them, rising to 95\% verification among the 354 recombinants that passed our QC filter. 
In addition to 3SEQ, we ran the CovRecomb~\cite{Li2024CovRecomb} 
and rebar recombination detection methods
on the putative recombinant sequences,
summarized in Table~\ref{tab:recombinant_validation}.
CovRecomb has a substantially lower validation rate
at 25\% among the QC-passing recombinants.
This is expected, as CovRecomb is specifically intended
to detect interlineage recombinants
and has previously been shown to be
less sensitive than other methods~\cite{Li2024CovRecomb}.
Rebar validation is intermediate, at 57\% among QC-passing recombinants.
(See also Figures~\ref{fig:recombinant_evidence_cov_recomb},
and \ref{fig:recombinant_evidence_rebar}.)
% We don't talk about RIPPLES results yet as we haven't discussed it, a
% and that would be confusing.

@@

Most of the deletions in these tables
occur in coding regions
(except the 1-base deletions at 28,271) and
do not cause frame shifts, involving 3, 6, or 9 sites
(except the 4-base deletion and the two 5-base deletions,
which are likely artefacts; see below).
All the deletions in these tables
(besides the artefactual cases)
have been linked to the emergence of major VOCs
(Alpha \cite{RambautAlpha2020},
Delta \cite{Stern2021},
Omicron B.1.1.529 \cite{RambautOmicron2021},
BA.1 \cite{RambautOmicron2021},
BA.2 \cite{RambautOmicron2021}, and
BA.4 \cite{Tegally2022})
and/or reported to be prevalent
in previous studies of deletions \cite{Li2023indels}.
All these major deletion events
(again, besides the artefactual cases)
are associated with the origins of major VOCs in the ARG
(see the Alpha subgraph in Figure~\ref{fig:subgraph_alpha};
the Delta subgraph in Figure~\ref{fig:subgraph_delta}; and
the Omicron subgraph in Figure~\ref{fig:subgraph_omicron}),
except the 1-base deletion at 28,271 on node 1436808
and the 9-base deletion at 686 on the BA.4-labeled node 2698748
(Table~\ref{tab:major_dels}).
Most of these deletions have been identified
as characteristic mutations of the major VOCs
in the studies or analysis cited in Table~\ref{tab:major_dels}.
Notably, eight of the major deletion events occur
in the N-terminal domain of the S1 unit of Spike
(a known hotspot of recurrent deletions \cite{McCarthy2021}
that plays a substantial role in antigenicity \cite{Harvey2021}),
which are all highly recurrent in the ARG
(Table~\ref{tab:rec_dels}).
These include known recurrent deletions:
the 6-base deletion at 21,765 and
the 3-base deletion at 21,991,
which are associated with
increased efficiency in cell entry \cite{Meng2021} and
antibody escape \cite{McCarthy2021}, respectively.
In addition, we found recurrent deletions in ORF8 and NSP3,
which have been suggested to be indel hotspots\cite{Akaishi2022}.
Together, these findings demonstrate that
the ARG captures genuine deletion events in SARS-CoV-2 evolution.
A comprehensive evolutionary analysis of the deletion events in the ARG
is beyond the scope of the current study;
however, we believe that the ARG, as a resource,
will enable and facilitate future studies
to better understand the role of deletions in SARS-CoV-2 evolution
(alongside recombination and single nucleotide mutations),
which to-date has been understudied in great part due to
a lack of suitable data models, methods, and tools.

@@

The B.1.1.7 (Alpha) variant emerged in the UK as the first major VOC.
Its origin is well reconstructed in the ARG,
with Figure \ref{fig:subgraph_alpha}
showing it to be a saltational lineage
characterized by a long branch preceding its origin.
Most of the characteristic Alpha mutations \cite{RambautAlpha2020}
(20 out of 23) in the ARG
occur immediately above the Alpha root.
However, a characteristic three-base correlated mutation
is instead associated with the parent node (labeled B.1.1) of the Alpha root,
suggesting that this mutation was acquired prior to the emergence of Alpha.
This parent node has an additional 5 children
(leading to 12 samples),
of which we display one (ERR4413600)
with no associated mutations,
meaning that it has an identical sequence
to the inferred ancestor of Alpha.
Below the root of Alpha, a single sample (ERR5178290)
appears without the characteristic mutation C3267T
(or any additional mutations).
However, this is only a single sample and
given that it also appears two months after the first Alpha samples
we treat its placement with caution.
Note that the list of mutations in ref\cite{RambautAlpha2020}
includes a T26801C in the M gene
which should no longer be considered characteristic of Alpha,
as the Wuhan-Hu-1 reference at 26,801 has since been updated to C.
In the ARG, a one-base non-coding deletion (position 28,271)
is also associated with the Alpha root;
although this is not in the original list of
characteristic Alpha mutations,
it has been suggested elsewhere as an Alpha-defining 
mutation\cite{Li2023indels}.

@@

Delta belongs to the ``B.1.617'' clade,
which also includes Kappa (B.1.617.1) and B.1.617.3 \cite{ECDCreport}.
Previous phylogenetic analyses suggest that
Delta and B.1.617.3 are more closely related to each other
than either is to Kappa\cite{Cherian2021,Farinholt2021}.
In the ARG, however, Delta and Kappa appear as sibling lineages
(see the subgraph in Figure~\ref{fig:subgraph_delta}).
This reflects the closest approximation to the relationships
suggested in the previous studies that could be inferred,
given that no B.1.617.3 samples passed our QC filters and
no seed B.1.617.3 samples were specified for ARG inference.

@@

\subsection{Additional analyses of the Pango X lineages}
\label{sec:supplement_indepth_pango_x_lineages}
In this section we provide some additional analysis on Pango X lineages
in the ARG. These analyses are intended to be read in conjunction with
the subgraph illustrations in Document S2.

\newcommand{\scghdiscussion}[1]{%
See \href{https://github.com/jeromekelleher/sc2ts-paper/discussions/#1}%
{github:sc2ts-paper/discussions/#1} for further details.}
\newcommand{\scghdiscussiontwo}[2]{%
See \href{https://github.com/jeromekelleher/sc2ts-paper/discussions/#1}%
{github:sc2ts-paper/discussions/#1} for further details on #2.}
\subsubsection{XA}
% mneumonic: deep dive XA
\label{sec:dd_xa}
XA traces to a well supported recombination node, with no reversions.
Note that the ``causal'' sample ERR5308556
(which triggers the initial recombination event)
is identical to the recombination node, and
therefore lacks C8090T,
which is one of the XA consensus mutations
(identified as those shared by over 90\% of XA samples).
As mentioned elsewhere,
the sc2ts detection results in terms of
the Pango lineages of the parents and the breakpoint intervals
agree with the results reported by Jackson et al. (Table \ref{tab:jackson}).
Two of the Jackson et al. XA samples
are present in the ARG (ERR5308556 and ERR5414941);
these samples are specifically marked in Figure~\ref{fig:subgraphs},
which is a summary of this subgraph.

@@

The displayed subgraph shows a small selection of the 192 XB samples.
XB was originally defined as a recombinant between
the closely related lineages B.1.634 (left parent) and B.1.631 (right parent)
from North and Central America \cite{Gutierrez2022}.
However, these parent lineages were not added to the ARG due to high HMM costs.
Together with other lineages such as B.1.627
(sister to B.1.631 \cite{Gutierrez2022}),
they are likely to attach somewhere along the long branch
in the subgraph visualization (with $> 30$ mutations).
If samples from these parental lineages were available for attachment,
then XB would be detected as a recombinant by sc2ts.
% which also contains the key ``R2'' deletion
% associated with XB / B.1.634 \cite{Gutierrez2022}.

% We hypothesize that there are a set of samples
% from a number of Pango lineages (including B.1.627, B.1.631, B.1.634)
% which, should they be added to the ARG, would break up the long branch,
% with XB being a recombination between two of them.
% In the absence of these, sc2ts cannot detect XB as a recombinant.
% Also, note that the XB consensus mutations all occur on or above the long branch.
% We suspect that the child node of the long branch
% represents the (potentially recombinant) origin of XB, and
% the B.1 samples at the bottom left of the subgraph
% may be misclassified by pangolin.

% Further work is required to determine
% why these additional samples were not gradually added
% to the ARG by the daily sample matching process or
% retrospective sample grouping.
